## NOAA Storm Database Analysis regarding Economy and Health Damages
#### Tianxiang(Ivan) Liu
#### 16/08/2014

### Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This analysis is mainly to address the questions including: 
    
    1. Across the United States, which types of events are most harmful with respect to population health? 
    
    2. Across the United States, which types of events have the greatest economic consequences? 

Main steps including data processing, data cleansing, plotting and resulting.

### Data processing
Download datasets and load into R
```{r, echo=TRUE}
setwd(choose.dir())
data.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
data.csv <- "repdata-data-StormData.csv"
if(!(file.exists(data.csv)))
    download.file(data.url, destfile="repdata-data-StormData.csv")
system.time(noaa.data <- read.csv(data.csv,head=T))
```

Dataset exploration - to get a summary re the datasets and allocate the variable names properly
```{r, echo=TRUE}
str(noaa.data)
head(noaa.data,n=3)
colnames(noaa.data)<-tolower(colnames(noaa.data))
```

Subset dataset based on analysis requirement - remove unnecessary variables in order to make dataset smaller and transfer damage figures to numbers with identical unit
```{r, echo=TRUE}
# first step subsetting
index.col <- c(1:2,7:8,21:28)
noaa.sub <- subset(noaa.data, select=index.col)
# calculate economic damage and health damage
noaa.sub$healthdmg <- noaa.sub$fatalities + noaa.sub$injuries
noaa.sub$cropmul[tolower(noaa.sub$cropdmgexp)=="k"]<-1000
noaa.sub$cropmul[tolower(noaa.sub$cropdmgexp)=="m"]<-1000000
noaa.sub$cropmul[is.na(noaa.sub$cropmul)] <- 1
noaa.sub$propmul[tolower(noaa.sub$propdmgexp)=="k"]<-1000
noaa.sub$propmul[tolower(noaa.sub$propdmgexp)=="m"]<-1000000
noaa.sub$propmul[is.na(noaa.sub$propmul)] <- 1
noaa.sub$cropdmg <- noaa.sub$cropdmg * noaa.sub$cropmul
noaa.sub$propdmg <- noaa.sub$propdmg * noaa.sub$propmul
noaa.sub$ecodmg <- noaa.sub$cropdmg + noaa.sub$propdmg
# second step subsetting
noaa.sub <- noaa.sub[,-c(1:3,5,6,10,12,14,15)]
```

Clean the inconsistencies in EVTYPE column - there is a significant percentage of event types are inconsistent such as "Avalanche" and "Avalanch", we need to indentify these errors and clean the data as much as possible. 
Data cleansing takes a lot of time and there are several steps in this section like first round cleasing and aggregation, second round cleasing and aggregation and mannual adjustments.
```{r,echo=TRUE}
event.types <- readLines("evtype.txt")
event.types <- as.factor(tolower(event.types))
noaa.sub$evtype2 <- as.factor(tolower(noaa.sub$evtype)) ## create new column for testing
spec_char <- c(" ","/","_",",",":","&","-")
for (i in 1:length(spec_char)){
    print(spec_char[i])
    event.types <- gsub(spec_char[i],"",as.character(event.types))
    noaa.sub$evtype2 <- gsub(spec_char[i],"",as.character(noaa.sub$evtype2))
}

## Matching event types based on event.types
for (i in 1:length(event.types)){
    index <- grepl(pattern=event.types[i],x=noaa.sub$evtype2, fixed=F)
    noaa.sub$eventtype[index] <- event.types[i]
}
## First aggregate to make datasets smaller for the efficiency of following cleansing steps
noaa.sub$eventtype[is.na(noaa.sub$eventtype)] <- noaa.sub$evtype2[is.na(noaa.sub$eventtype)]
library(plyr)
noaa.agg <- ddply(noaa.sub, .(eventtype), summarize, propdmg=sum(propdmg),cropdmg=sum(cropdmg),
                  fatalities=sum(fatalities),injuries=sum(injuries),healthdmg=sum(healthdmg),
                  ecodmg=sum(ecodmg))
noaa.agg$sum <- noaa.agg[,6]+noaa.agg[,7]
noaa.agg <- noaa.agg[!(noaa.agg$sum==0),]
noaa.agg <- noaa.agg[,-8]
## Mannual data cleansing
adj <- matrix(c("tstm","thunderstormwind","thunder","thunderstormwind",
                "thuder","thunderstormwind","avalan","avalanche",
                "astronomical","astronomicallowtide","fog","freezingfog",
                "hurricane","hurricane(typhoon)","cold","coldwindchill",
                "torndao","tornado","thunerstormwinds","thunderstormwind",
                "thundeerstormwinds","thunderstormwind",
                "freeze","frost/freeze"), ncol=2,byrow = T)
for (i in 1:length(adj[,1])){
    for (j in 1:length(noaa.agg$eventtype)){
        if(grepl(adj[i,1],noaa.agg$eventtype[j],fixed=F))
            noaa.agg$eventtype[j] <- adj[i,2]
    }
}
noaa.agg <- ddply(noaa.agg, .(eventtype), summarize, propdmg=sum(propdmg),cropdmg=sum(cropdmg),
                  fatalities=sum(fatalities),injuries=sum(injuries),healthdmg=sum(healthdmg),
                  ecodmg=sum(ecodmg))
```

Sort and subset datasets for different analysis - only 10 largest items are subsetted into dataset
```{r,echo=TRUE}
noaa.health <- noaa.agg[,-c(2,3,7)]
noaa.eco <- noaa.agg[,-c(4,5,6)]
noaa.health <- head(arrange(noaa.health,desc(healthdmg)),n=10)
noaa.eco <- head(arrange(noaa.eco,desc(ecodmg)),n=10)
```

### Results
Health Plots and Results
```{r,echo=TRUE,fig.height=6,fig.width=12}
library(ggplot2)
g1 <- ggplot(noaa.health,aes(x=eventtype, y=fatalities))
g1 + geom_bar(stat="identity", fill="red",alpha=1, colour="black",) + 
labs(x="Health damage") + labs(y="Event types") + 
labs(title=paste("Most Harmful Events to Population Health", "(Top 10)", sep="\n"))+
theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
geom_bar(aes(x=eventtype, y=injuries),stat="identity",fill="red",alpha=0.3, colour="black") +
geom_bar(aes(x=eventtype, y=healthdmg),stat="identity",fill="red",alpha=0.3, colour="black")
```

Economy Plots and Results
```{r,echo=TRUE, fig.height=6,fig.width=12}
g2 <- ggplot(noaa.eco,aes(x=eventtype, y=propdmg))
g2 + geom_bar(stat="identity", fill="black",alpha=0.5, colour="black",) + 
labs(x="Health damage") + labs(y="Event types") + 
labs(title=paste("Most Harmful Events to Economy", "(Top 10)", sep="\n"))+
theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
geom_bar(aes(x=eventtype, y=cropdmg),stat="identity",fill="blue",alpha=0.5, colour="black") +
geom_bar(aes(x=eventtype, y=ecodmg),stat="identity",fill="blue",alpha=0.5, colour="black")    
```